---
title: "Final Project"
author: "David Adeyemo"
date: "2/21/2022"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r}
pkg <- c(
  "janitor", "here", "tidyverse", "readr", "ggplot2", "knitr",
  "mlr3", "plotly", "data.table", "tidyr", "shiny", "xgboost", "caret")

install.packages(pkg)
```


```{r}
library(tidyverse)
library(readr)
library(janitor)
library(here)
```

# The 2010 Data

```{r}
read.csv(
  here("Data", "FARS2010/ACCIDENT.CSV"), sep = ",",
  header = TRUE
  ) %>%
  clean_names() -> accident_2010
save(accident_2010, file = "Data/FARS2010/accident_2010.RData")

read.csv(
  here("Data", "FARS2010/VEHICLE.CSV"), sep = ",",
  header = TRUE
  ) %>%
  clean_names() -> vehicle_2010
save(vehicle_2010, file = "Data/FARS2010/vehicle_2010.RData")

read.csv(
  here("Data", "FARS2010/PERSON.CSV"), sep = ",",
  header = TRUE
  ) %>%
  clean_names() -> person_2010 
save(person_2010, file = "Data/FARS2010/person_2010.RData")

read.csv(
  here("Data", "FARS2010/ACC_AUX.CSV"), sep = ",",
  header = TRUE
  ) %>%
  clean_names() -> acc_aux_2010 
save(acc_aux_2010, file = "Data/FARS2010/acc_aux_2010.RData")

read.csv(
  here("Data", "FARS2010/CEVENT.CSV"), sep = ",",
  header = TRUE
  ) %>%
  clean_names() -> cevent_2010 
save(cevent_2010, file = "Data/FARS2010/cevent_2010.RData")

read.csv(
  here("Data", "FARS2010/DISTRACT.CSV"), sep = ",",
  header = TRUE
  ) %>%
  clean_names() -> distract_2010 
save(distract_2010, file = "Data/FARS2010/distract_2010.RData")

read.csv(
  here("Data", "FARS2010/DRIMPAIR.CSV"), sep = ",",
  header = TRUE
  ) %>%
  clean_names() -> drimpair_2010 
save(drimpair_2010, file = "Data/FARS2010/drimpair_2010.RData")

read.csv(
  here("Data", "FARS2010/FACTOR.CSV"), sep = ",",
  header = TRUE
  ) %>%
  clean_names() -> factor_2010 
save(factor_2010, file = "Data/FARS2010/factor_2010.RData")

read.csv(
  here("Data", "FARS2010/MANEUVER.CSV"), sep = ",",
  header = TRUE
  ) %>%
  clean_names() -> maneuver_2010 
save(maneuver_2010, file = "Data/FARS2010/maneuver_2010.RData")

read.csv(
  here("Data", "FARS2010/MIACC.CSV"), sep = ",",
  header = TRUE
  ) %>%
  clean_names() -> miacc_2010 
save(miacc_2010, file = "Data/FARS2010/miacc_2010.RData")

read.csv(
  here("Data", "FARS2010/MIDRVACC.CSV"), sep = ",",
  header = TRUE
  ) %>%
  clean_names() -> midrvacc_2010 
save(midrvacc_2010, file = "Data/FARS2010/midrvacc_2010.RData")

read.csv(
  here("Data", "FARS2010/MIPER.CSV"), sep = ",",
  header = TRUE
  ) %>%
  clean_names() -> miper_2010 
save(miper_2010, file = "Data/FARS2010/miper_2010.RData")

read.csv(
  here("Data", "FARS2010/NMCRASH.CSV"), sep = ",",
  header = TRUE
  ) %>%
  clean_names() -> nmcrash_2010 
save(nmcrash_2010, file = "Data/FARS2010/nmcrash_2010.RData")

read.csv(
  here("Data", "FARS2010/NMIMPAIR.CSV"), sep = ",",
  header = TRUE
  ) %>%
  clean_names() -> nmimpair_2010 
save(nmimpair_2010, file = "Data/FARS2010/nmimpair_2010.RData")

read.csv(
  here("Data", "FARS2010/NMPRIOR.CSV"), sep = ",",
  header = TRUE
  ) %>%
  clean_names() -> nmprior_2010 
save(nmprior_2010, file = "Data/FARS2010/nmprior_2010.RData")

read.csv(
  here("Data", "FARS2010/PARKWORK.CSV"), sep = ",",
  header = TRUE
  ) %>%
  clean_names() -> parkwork_2010 
save(parkwork_2010, file = "Data/FARS2010/parkwork_2010.RData")

read.csv(
  here("Data", "FARS2010/PER_AUX.CSV"), sep = ",",
  header = TRUE
  ) %>%
  clean_names() -> per_aux_2010 
save(per_aux_2010, file = "Data/FARS2010/per_aux_2010.RData")

read.csv(
  here("Data", "FARS2010/SAFETYEQ.CSV"), sep = ",",
  header = TRUE
  ) %>%
  clean_names() -> safetyeq_2010 
save(safetyeq_2010, file = "Data/FARS2010/safetyeq_2010.RData")

read.csv(
  here("Data", "FARS2010/VEH_AUX.CSV"), sep = ",",
  header = TRUE
  ) %>%
  clean_names() -> veh_aux_2010 
save(veh_aux_2010, file = "Data/FARS2010/veh_aux_2010.RData")

read.csv(
  here("Data", "FARS2010/VEVENT.CSV"), sep = ",",
  header = TRUE
  ) %>%
  clean_names() -> vevent_2010 
save(vevent_2010, file = "Data/FARS2010/vevent_2010.RData")

read.csv(
  here("Data", "FARS2010/VIOLATN.CSV"), sep = ",",
  header = TRUE
  ) %>%
  clean_names() -> violatn_2010 
save(violatn_2010, file = "Data/FARS2010/violatn_2010.RData")

read.csv(
  here("Data", "FARS2010/VISION.CSV"), sep = ",",
  header = TRUE
  ) %>%
  clean_names() -> vision_2010 
save(vision_2010, file = "Data/FARS2010/vision_2010.RData")

read.csv(
  here("Data", "FARS2010/VSOE.CSV"), sep = ",",
  header = TRUE
  ) %>%
  clean_names() -> vsoe_2010 
save(vsoe_2010, file = "Data/FARS2010/vsoe_2010.RData")
```

Now to merge all the files for 2010. But first, how can we figure out what column names are identical in the dataframes, since that will have to be used to merge the files?

```{r}
intersect(
  colnames(acc_aux_2010), 
  colnames(accident_2010)
  )

ac_a2010 <- acc_aux_2010 %>% 
  left_join(
    accident_2010,
    by = c("year", "state", "st_case", "county", "fatals")
  )
```
```{r}
intersect(
  colnames(ac_a2010),
  colnames(cevent_2010)
)

ac_a2010 <- ac_a2010 %>% 
  left_join(
    cevent_2010,
    by = c("state", "st_case")
  )
```
```{r}
intersect(
  colnames(ac_a2010),
  colnames(distract_2010)
)

ac_a2010 <- ac_a2010 %>% 
  left_join(
    distract_2010,
    by = c("state", "st_case")
  )
```
```{r}
intersect(
  colnames(ac_a2010),
  colnames(drimpair_2010)
)

ac_a2010 <- ac_a2010 %>% 
  left_join(
    drimpair_2010,
    by = c("state", "st_case", "veh_no")
  )
```
```{r}
intersect(
  colnames(ac_a2010),
  colnames(factor_2010)
)

ac_a2010 <- ac_a2010 %>% 
  left_join(
    factor_2010,
    by = c("state", "st_case", "veh_no")
  )
```
```{r}
intersect(
  colnames(ac_a2010),
  colnames(maneuver_2010)
)

ac_a2010 <- ac_a2010 %>% 
  left_join(
    maneuver_2010,
    by = c("state", "st_case")
  )
```
```{r}
intersect(
  colnames(ac_a2010),
  colnames(miacc_2010)
)

ac_a2010 <- ac_a2010 %>% 
  left_join(
    miacc_2010,
    by = c("st_case")
  )
```
```{r}
intersect(
  colnames(ac_a2010),
  colnames(midrvacc_2010)
)

ac_a2010 <- ac_a2010 %>% 
  left_join(
    midrvacc_2010,
    by = c("st_case", "a1", "a2", "a3",
           "a4", "a5", "a6", "a7", "a8", "a9", "a10")
  )
```
```{r}
intersect(
  colnames(ac_a2010),
  colnames(miper_2010)
)

ac_a2010 <- ac_a2010 %>% 
  left_join(
    miper_2010,
    by = c("st_case")
  )
```
```{r}
intersect(
  colnames(ac_a2010),
  colnames(nmcrash_2010)
)

ac_a2010 <- ac_a2010 %>% 
  left_join(
    nmcrash_2010,
    by = c("state", "st_case", "veh_no", "per_no")
  )
```
```{r}
intersect(
  colnames(ac_a2010),
  colnames(nmimpair_2010)
)

ac_a2010 <- ac_a2010 %>% 
  left_join(
    nmimpair_2010,
    by = c("state", "st_case", "veh_no", "per_no")
  )
```
```{r}
intersect(
  colnames(ac_a2010),
  colnames(nmprior_2010)
)

ac_a2010 <- ac_a2010 %>% 
  left_join(
    nmprior_2010,
    by = c("state", "st_case", "veh_no", "per_no")
  )
```
```{r}
intersect(
  colnames(ac_a2010),
  colnames(parkwork_2010)
)

ac_a2010 <- ac_a2010 %>% 
  left_join(
    parkwork_2010,
    by = c("state", "st_case", "veh_no")
  )
```
```{r}
intersect(
  colnames(ac_a2010),
  colnames(per_aux_2010)
)

ac_a2010 <- ac_a2010 %>% 
  left_join(
    per_aux_2010,
    by = c("year", "st_case", "veh_no", "per_no")
  )
```
```{r}
intersect(
  colnames(ac_a2010),
  colnames(person_2010)
)

ac_a2010 <- ac_a2010 %>% 
  left_join(
    person_2010,
    by = c("state", "st_case", "county", "ve_forms", "day",
           "month", "hour", "minute", "road_fnc", "harm_ev",
           "man_coll", "sch_bus", "veh_no", "per_no")
  )
```
```{r}
intersect(
  colnames(ac_a2010),
  colnames(safetyeq_2010)
)

ac_a2010 <- ac_a2010 %>% 
  left_join(
    safetyeq_2010,
    by = c("state", "st_case", "veh_no", "per_no")
  )
```
```{r}
intersect(
  colnames(ac_a2010),
  colnames(veh_aux_2010)
)

ac_a2010 <- ac_a2010 %>% 
  left_join(
    veh_aux_2010,
    by = c("year", "st_case", "veh_no")
  )
```
```{r}
intersect(
  colnames(ac_a2010),
  colnames(vehicle_2010)
)

ac_a2010 <- ac_a2010 %>% 
  left_join(
    vehicle_2010,
    by = c("state", "st_case", "ve_forms", "day", "month", "hour",
           "minute", "harm_ev", "man_coll", "veh_no", "make", "mak_mod",
           "body_typ", "mod_year", "vintype", "vinmake", "vina_mod",
           "vin_bt", "vinmodyr", "vin_lngt", "vin_wgt", "wgtcd_tr",
           "whlbs_lg", "whlbs_sh", "ser_tr", "fuelcode", "mcycl_ds",
           "tow_veh", "spec_use", "emer_use", "rollover", "impact1",
           "impact2", "fire_exp"
           )
  )
```
```{r}
intersect(
  colnames(ac_a2010),
  colnames(vevent_2010)
)

ac_a2010 <- ac_a2010 %>% 
  left_join(
    vevent_2010,
    by = c("state", "st_case", "eventnum", "vnumber1", "aoi1", "soe",
           "vnumber2", "aoi2", "veh_no")
  )
```
```{r}
intersect(
  colnames(ac_a2010),
  colnames(violatn_2010)
)

ac_a2010 <- ac_a2010 %>% 
  left_join(
    violatn_2010,
    by = c("state", "st_case", "veh_no")
  )
```

```{r}
intersect(
  colnames(ac_a2010),
  colnames(vision_2010)
)

ac_a2010 <- ac_a2010 %>% 
  left_join(
    vision_2010,
    by = c("state", "st_case", "veh_no")
  )
```
```{r}
intersect(
  colnames(ac_a2010),
  colnames(vsoe_2010)
)

ac_a2010 <- ac_a2010 %>% 
  left_join(
    vsoe_2010,
    by = c("state", "st_case", "soe", "veh_no", "veventnum")
  )
```

ac_a2010 is a combination of all the data files in 2010

```{r}
save(ac_a2010, file = "Data/ac_a2010.RData")
```

# Just the accident, person, vehicle file

```{r}
intersect(
  colnames(accident_2010), 
  colnames(person_2010)
  )

accident_2010 %>%
  left_join(
    person_2010, 
    by = c(
      "state", "st_case", "ve_forms", "county", "day",
      "month", "hour", "minute", "road_fnc", "harm_ev",
      "man_coll", "sch_bus"
      )
    ) -> ap2010 

intersect(
  colnames(ap2010), 
  colnames(vehicle_2010)
  )

ap2010 %>%
  left_join(
    vehicle_2010,
    by = c(
      "state", "st_case", "ve_forms", "day", "month", "hour",
      "minute", "harm_ev", "man_coll", "veh_no", "make",
      "mak_mod", "body_typ", "mod_year", "vintype", "vinmake",
      "vina_mod", "vin_bt", "vinmodyr", "vin_lngt", "vin_wgt",
      "wgtcd_tr", "whlbs_lg", "whlbs_sh", "ser_tr", "fuelcode",
      "mcycl_ds", "tow_veh", "spec_use", "emer_use", "rollover",
      "impact1", "impact2", "fire_exp"
      )
    ) -> apv_2010

save(apv_2010, file = "Data/apv_2010.RData")

```

Let us add the state abbreviations and state names. We can use the `states` dataframe in the `urbnmapr` package. But we will have to convert our state variable into state_fips first, and have it be in the character format.

```{r}
library(urbnmapr)
data(states)
names(states)

states %>%
  distinct(
    state_fips, state_abbv, state_name
    ) -> stdf

apv_2010 %>%
  mutate(
    state_fips = stringr::str_pad(
      state, width = 2, side = "left", pad = "0")
    ) %>%
  relocate(state, state_fips, state_abbv, state_name) %>%
  select(-state) -> apv_2010

apv_2010 %>%
  left_join(
    stdf,
    by = "state_fips"
    ) -> apv_2010
```

Now, what was the date and time of the accident? Well, to construct that detail we will have to create a new `date_time` variable in two pieces, first the date, and then the time. 

```{r}
apv_2010 %>%
  mutate(
    date = paste(year, month, day, sep = "-"),
    time = paste(hour, minute, 0, sep = "-")
    ) %>%
  relocate(
    year, month, day, date, hour, minute, time
    ) -> apv_2010
```

Now we will combine date and time, and then format it as a date-time field so R will know what it is and no treat it as a number or a character string. 

```{r}
apv_2010 %>%
  mutate(
    date_time = paste(date, time, sep = ":"),
    date_time = lubridate::ymd_hms(date_time)
    ) %>%
  relocate(date_time) -> apv_2010
```

Now we can drop the original columns since `date_time` contains all the information in a single column.

```{r}
apv_2010 %>%
  select(
    -c(year, month, day, date, hour, minute, time)
    ) -> apv_2010
```

What about labeling the `day_week`?

```{r}
apv_2010 %>%
  mutate(
    day_of_week = lubridate::wday(
      date_time, 
      abbr = FALSE, 
      label = TRUE
      )
    ) %>%
  relocate(
    date_time, day_of_week
    ) %>%
  select(
    -c(day_week)
    ) -> apv_2010
```

Labeling some of the other variables?

```{r}
apv_2010 %>%
  mutate(
    route = case_when(
      route == 1 ~ "Interstate",
      route == 2 ~ "U.S. Highway",
      route == 3 ~ "State Highway",
      route == 4 ~ "County Highway",
      route == 5 ~ "Local Street - Township",
      route == 6 ~ "Local Street - Municipality",
      route == 7 ~ "Local Street - Frontage",
      route == 8 ~ "Other",
      route == 9 ~ "Unknown"
      ),
    sch_bus = case_when(
      sch_bus == 0 ~ "School Bus Not Involved",
      sch_bus == 1 ~ "School Bus Involved"
      )
    ) -> apv_2010 

save(apv_2010, file = "Data/apv_2010.RData")
```

Now the rest of the variables would be labeled as needed. 

Of course, you could wait until you have all the yearly files combined into a single one and then label the variables of interest. 

# The 2019 Data

```{r}
read.csv(
  here("Data", "FARS2019/ACCIDENT.CSV"), sep = ",",
  header = TRUE
  ) %>%
  clean_names() -> accident_2019

read.csv(
  here("Data", "FARS2019/VEHICLE.CSV"), sep = ",",
  header = TRUE
  ) %>%
  clean_names() -> vehicle_2019

read.csv(
  here("Data", "FARS2019/PERSON.CSV"), sep = ",",
  header = TRUE
  ) %>%
  clean_names() -> person_2019

```








